Testing and Isolation Efficacy; Insights from a Simple Epidemic Model

### 2021-01-05 SIR_manuscript.tex
intro updated

The observed dynamics of the COVID-19 epidemic is dirven by the underlying processes including both epidemiological processes (infection, recovery, etc.) and testing processes (who gets tested and intensity). In addition to determining what is actually observed (case reports), testing processes can also feed back to affect the epidemiological dynamics through isolation and quarantine. Specifically, individuals who are in the process of testing may partially or fully self-isolate, and individuals who have tested positive are highly likely to self-isolate. We developed a model that incorporates mechanistic epidemic processes and testing to explore the effects of testing and isolation on the epidemic dynamics, and providing insights for the population-level epidemic control.

During COVID-19 epidemic, since isolation has applied to people including, but not limited to, those who have been tested, the epidemic dynamics will depend on who gets tested. That is, how tests are distributed across epidemiological compartments for a given testing intensity. Specifically, in the case where tests are randomly assigned within the population, the tests will be weighted equally across all compartments. Random testing is the simplest form of testing regime and might happen, for example, when disease surveillance is the purpose of testing. However, in the case where targeted testing is administered within the population, the testing will be more weighted towards the compartments with people who have higher probability of getting tested. Targeted testing might happen, for example, for different purposes including diagnosis and treatment, screening people for access to things like flights, surgery, long-term-care facilities, and for mitigation and control of the disease spread. Specifically, when the purpose is "diagnosis and treatment" or "mitigation and control of disease spread", the bias is skewed towards infectious individuals with even more weights for the diagnosis and treatment than the mitigation and control. Also, when people are tested for "screening", the bias is towards the group-specific characteristics \ali{better wording needed}. For example, people who are getting on flights may or will be more mobile than people awaiting for a surgery and are probably going to stay in a long-term care facility and consequently more isolated to begin with.

Furthermore, test-specific sensitivities, intensity and sample-to-answer reporting time are some components of the testing processes that influence the epidemic dynamic. We model these components and explore the trade-offs between them in achieving the epidemic control objective.

R0

Frequent proactive testing of asymptomatic individuals offers an additional way to reduce transmission, and is among the most powerful and least disruptive of non-pharmaceutical interventions such as face masks, social distancing.

% \bmb{As JD suggests, the focus of the paper is not on unraveling data (although that is something we \emph{could} do, and are trying to do, with this modeling framework as well). I would say that the ``bullet points'' to start the paper are (1) the processes driving the observed dynamics of the COVID-19 epidemic include both epidemiological processes (infection, recovery, etc.) and testing processes (``processes'' is repeated here, could rephrase); (2) in addition to determining what is actually observed (case reports), testing processes can also feed back to affect the epidemiological dynamics through isolation and quarantine; (3) your last sentence in the \P\ above.}

\ali{discussion maybe: all of the accessible testing data is being used for one of the other three purposes rather than surveillance. These are targeted testing strategies with a different bias.}

- Speed of test; 
Much effort has been invested in developing widely accessible and fast tests in order to mitigate the spread of the pandemic (refs). For example, PCR tests are commonly used, and very sensitive, but take longer compared to more recent tests, such as a breath-based test currently in trial stage, which can provide a result in seconds.

@@

>>>>

The observed dynamics of the COVID-19 epidemic is dirven by the underlying processes including both epidemiological processes (infection, recovery, etc.) and testing processes (including who gets tested for a given testing intensity, test sensitivity, etc. \ali{right?}). In addition to determining what is actually observed (case reports), testing processes can also feed back to affect the epidemiological dynamics through isolation and quarantine. Specifically, individuals who are in the process of testing may partially or fully self-isolate, and individuals who have tested positive are highly likely to self-isolate. We developed a model that incorporates mechanistic epidemic processes and testing to explore the effects of testing and isolation on the epidemic dynamics, and providing insights for the population-level epidemic control.

During COVID-19 epidemic, since isolation has applied to people including, but not limited to, those who have been tested, the epidemic dynamics will depend on who gets tested. That is, how tests are distributed across epidemiological compartments (susceptible-infectious-recovered) for a given testing intensity. Specifically, in the case where tests are randomly assigned within the population, the tests will be weighted equally across all compartments. Random testing strategy is the simplest form of testing and might happen, for example, when disease surveillance is the purpose of testing \citep{foddai2020base}. However, in the case where targeted testing is administered within the population, the testing will be more weighted towards the compartments with people who have higher probability of getting tested. Targeted testing might happen, for example, for different purposes including diagnosis and treatment, screening people for access to things like flights, surgery, long-term-care facilities, and for the purpose of mitigation and control of the disease spread. Specifically, when the purpose is "diagnosis and treatment" or "mitigation and control of disease spread", the weights of testing is skewed towards infectious compartment rather than susceptible and recovered compartments. Further, heavier testing weight may be put on the infected compartment in the case of diagnosis and treatment than the case of mitigation and control \ali{why? refs?}. Also, when people are tested for "screening", the individuals with potential higher mobility, eg. people who are getting on flights, get more tested and thus the coresponding heavier testing weight is assigned than people awaiting for a surgery and are probably going to stay in a long-term care facility and consequently less mobile and more isolated to begin with.

In addition to the efect of compartment-specific testing weights on the isolation, and thus epidemic dynamics, other influential components of testing processes on isolation include test-specific sensitivities, intensity and sample-to-answer reporting time. They determine how accurately and how fast people are put in the isolation. We model these components and explore their effects on the epidemic dynamics. Specifically, we study the trade-offs between the testing strategies (random and targeted) and the discussed testing components on a single epidemiological-dynamics index, the ``basic reproduction number'' $\Rnum$.

The early dynamics of an epidemic are determined by $\Rnum$. This index is defined as the expected number of secondary infections arising from a typical infective individual in a completely susceptible population \citep{dietz1993estimation}. In the early stages of an epidemic the number of infected individuals is expected grow exponentially over time when $\Rnum>1$. To control the spread of disease in the population, $\Rnum$ is required to be below unity.

@@

% Frequent proactive testing of asymptomatic individuals offers an additional way to reduce transmission, and is among the most powerful and least disruptive of non-pharmaceutical interventions such as face masks, social distancing.
% \bmb{As JD suggests, the focus of the paper is not on unraveling data (although that is something we \emph{could} do, and are trying to do, with this modeling framework as well). I would say that the ``bullet points'' to start the paper are (1) the processes driving the observed dynamics of the COVID-19 epidemic include both epidemiological processes (infection, recovery, etc.) and testing processes (``processes'' is repeated here, could rephrase); (2) in addition to determining what is actually observed (case reports), testing processes can also feed back to affect the epidemiological dynamics through isolation and quarantine; (3) your last sentence in the \P\ above.}

% - Speed of test; 
% Much effort has been invested in developing widely accessible and fast tests in order to mitigate the spread of the pandemic (refs). For example, PCR tests are commonly used, and very sensitive, but take longer compared to more recent tests, such as a breath-based test currently in trial stage, which can provide a result in seconds.
\ali{discussion maybe: all of the accessible testing data is being used for one of the other three purposes rather than surveillance. These are targeted testing strategies with a different bias.}
